<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{Pretty and parallel heatmap}
-->

Pretty and parallel heatmap
===========================

**Author**: Zuguang Gu ( z.gu@dkfz.de )

First, let's generate a matrix:

```{r}
set.seed(123)
mat = rbind(matrix(rnorm(50, 1), 5, ),
	        matrix(rnorm(50, -1), 5, 10))
colnames(mat) = paste0("c", 1:10)
rownames(mat) = paste0("r", 1:10)
```

plot the matrix:

```{r, message = FALSE}
library(pheatmap2)
pheatmap2(mat)
pheatmap2(mat, cluster_rows = FALSE, cluster_cols = FALSE, 
    show_colnames = FALSE, show_rownames = FALSE,
    legend_title = NULL)
library(circlize)
pheatmap2(mat, col_fun = colorRamp2(c(-3, 0, 3), c("green", "white", "red")))
```

If `col_fun` is set, the color mapping is resistant to outliers.

```{r}
mat2 = mat
mat2[1, 1] = 10
mat2[1, 2] = -10
pheatmap2(mat2, col_fun = colorRamp2(c(-3, 0, 3), c("green", "white", "red")))
```

Define column annotations and corresponding colors

```{r}
annotation = data.frame(label1 = sample(c("a", "b"), 10, replace = TRUE),
                        label2 = sample(c("c", "d"), 10, replace = TRUE),
						label3 = 1:10)
rownames(annotation) = colnames(mat)
annotation

annotation_colors = list(label1 = c("a" = "red", "b" = "blue"),
	                     label2 = c("c" = "yellow", "d" = "green"))
annotation_colors
```

plot the heatmap with annotations:

```{r}
pheatmap2(mat, annotation = annotation)
pheatmap2(mat, annotation = annotation, annotation_colors = annotation_colors)
```

generate an additional matrix

```{r}
mat2 = matrix(runif(40), 10, 4)
colnames(mat2) = paste0("d", 1:4)
```

plot the heatmap with the additional matrix

```{r fig.width = 10, fig.height = 7}
pheatmap2(mat, 
	legend_title = "mat", 

	mat_list = list(mat2 = mat2),
	col_fun_list = list(mat2 = colorRamp2(c(0, 0.5, 1), c("green", "white", "red"))),
	legend_title_list = "mat2")
```

Set gaps between heatmap and the title/sub titles

```{r fig.width = 10, fig.height = 7}
pheatmap2(mat, 
	legend_title = "mat", 
	main = "two heatmap", sub_main = c("mat1", "mat2"),
	gap = 3,

	mat_list = list(mat2 = mat2),
	col_fun_list = list(mat2 = colorRamp2(c(0, 0.5, 1), c("green", "white", "red"))),
	legend_title_list = "mat2")
```

Annotations for the additional matrix

```{r}
annotation2 = data.frame(label1 = c("a", "a", "b", "b"), label5 = 1:4)
rownames(annotation2) = colnames(mat2)
annotation_colors2 = list(label1 = c("a" = "red", "b" = "blue"))
```

The complex heatmap

```{r fig.width = 10, fig.height = 7}
pheatmap2(mat,
	legend_title = "mat", 
	annotation = annotation, 
	annotation_colors = annotation_colors,
	
	mat_list = list(mat2 = mat2),
	col_fun_list = list(mat2 = colorRamp2(c(0, 0.5, 1), c("green", "white", "red"))),
	legend_title_list = "mat2",
	annotation_list = list("mat2" = annotation2),
	annotation_colors_list = list("mat2" = annotation_colors2))
```						 

Column annotation can also be character.

```{r}
group = matrix(ifelse(rowMeans(mat) > 0, "high", "low"), ncol = 1)
colnames(group) = "group"
```

plot the heatmap with the additional matrix

```{r fig.width = 10, fig.height = 7}
pheatmap2(mat, 
	legend_title = "mat", 

	mat_list = list(group = group),	
	col_fun_list = list(group = colorRamp2(seq_along(unique(group)), c("red", "blue"))),
	legend_title_list = "group")
```

Finally, let's make a heatmap which contains everything.

```{r fig.width = 10, fig.height = 7}
pheatmap2(mat, 
	legend_title = "mat", 
	annotation = annotation, 
	annotation_colors = annotation_colors,
	
	mat_list = list(group = group, mat2 = mat2),
	col_fun_list = list(group = colorRamp2(seq_along(unique(group)), c("red", "blue")),
                        mat2 = colorRamp2(c(0, 0.5, 1), c("green", "white", "red"))),
	legend_title_list = c("group", "mat2"),
	annotation_list = list("mat2" = annotation2),
	annotation_colors_list = list("mat2" = annotation_colors2))
```	
